Child survival for mothers: mortality change and related measures

### 2020-08-11 paper.Rmd
Aclaraciones,  mejora redaccion

Kin count estimation has a long and distinguished pedigree in mathematical demography, starting with the work of @lotka1931orphanhood on modelling orphanhood in theoretical populations across demographic regimes. To the best of our knowledge, @brass_derivation_1953 first proposed a formula to represent child survival over age. The notion of estimating the expected number of living daughters in a stable population was generalized by @goodman1974 for other kin relations (granddaughters, cousins, etc.). The "counting method" approach was further popularized in @KeyfitzMath. @bongaarts_projection_1987 used a similar approach to estimate descendants in his 'Family Status Model'.

Child survival - and it's counterpart, child death - gain new significance in the wake of the fertility transition.
Depressed fertility means that fewer children are tasked with providing key emotional, social, and financial transfers to aging parents for ever increasing periods of time.
With increasing periods of generational overlap, individuals find themselves 'sandwiched' between aging parents and young children requiring their simultaneous attention and care [@daatland_norwegian_2010].
In the context of global population aging, elderly parents without access to formal social security and pension systems are particularly reliant on these transfers to make ends meet [@smith-greenaway_maternal_2020].  
Losing an only child may become a more common experience in the context of global fertility decline, something particularly worrying for parents suddenly rendered childless.
Given the known psychological, health, and social consequences of child loss for bereaved parents and families [@Hendrickson2009;@lee_death_2014], it is surprising that the demographic processes that shape parental bereavement remain very poorly understood.
Recent development have underscored the need to understand how sudden changes in mortality affect the availability of kin. 
In the context of the global pandemic caused by the Covid-19 disease, elderly people depend on support provided by their relatives, especially during periods of lockdown and in the absence of governmental support mechanisms.<!-- Maybe here are repeated --> Elderly parents may be at a higher risk of losing the key support provided by their adult children given the known age-gradient in the Covid-19 case fatality rate. 
How do changes in mortality affect the availability of offspring over age from the point of view of a prospective mother? How can we characterize the timing of offspring mortality as experienced by a mother from a formal demographic perspective? How this would work in an heterogeneity population?
In the remainder of the paper, we aim to formalize the relationship population-level changes in mortality rates and perceived change in the lived experience of child death from the perspective of a mother. Our approach is key for understanding the impact of age-specific excess mortality on the resilience or otherwise of kinship networks for the vulnerable elderly population.
We consider the case when these changes are additive and mutliplicative. In this paper, we uncover formal relationships to answer these questions and provide numerical applications using data from countries in Latin America during the second half of the twentieth century. See the [Applications](#Applications) section for more details on the source, calculation procedures and main numerical results. We used period measures, in an stable context, so the objetive is give moment measures. A more sophisticated model using the appropriate age-specific cohort rates in the subscripts has been proposed by @alburez_gutierrez_womens_2019.

@@

As an initial step, we seek an intuitive understanding of the effect of changing mortality on child survival. 
Building on work by @KeyfitzMath for the probability of a living mother, we can show which features of the child survival function explain the expected number of children. 
For this we use an approximate of $l_x$ using Taylor´s theorem until second order around the mean age of childbearing $\kappa$:

@@

We find that, seen from the perspective of a mother, child survival mainly depends on the cummulative fertility function^[Different from the findings of @KeyfitzMath of having living ancestors.] and the survival from birth to $a-\kappa$. Also that the approximation is affected (negativally) by how disperse is the fertility by age (variance $\sigma^2$) and the suvival curvature around (with negative sign), in an age range where $l_x$ is tipycally very flat (range 20-40 years old) in transitioned populations^[This approximation is useful also for get a dimension changes in the fertility average age \kappa: taking logs in $F_a \, l_{a-\kappa}$ and deriving we get $\frac{\Delta CS_a}{CS_a} \approx - \mu_{a-\kappa} \Delta\kappa$].

@@

$$\frac{dCS^{\delta}}{d\delta} = -a\int_{\alpha}^{a}m_{x}{l_{a-x} dx} + \int_{\alpha}^{a} m_{x} {l_{a-x}x dx}\\
= -a \, CS_{a}  + \int_{\alpha}^{a} {x  \, m_{x} l_{a-x} dx}$$

@@

The expected change in descendants survival is inversely proportional to the difference between maternal age $a$ and the mean age of the mother at the birth of her surviving daughters $k_a$. The magnitude of the change depends negatively on the age distribution of the surviving offspring. This is intuitive considering that younger descendants experiences longer periods of exposure to risk.

That an equal change in mortality could happens at all ages is unlikelihood. This measure could be much related a change in first ages (child and youth) than adult or ones. As a corollary, if the change only happens in the age range $\left[0;1\right)$, infant mortality, we can inspect the effect starting with splitting the integral (considering those born last year and not, and of course for $a>1$):

@@

$$CS_{a}^{\delta_0} =-\int_{0}^{a-1} m_{x} l_{a-x} dx - a \int_{a-1}^{a} {m_{x} l_{a-x} dx} + \int_{a-1}^{a} {x \, m_{x} l_{a-x} dx}$$
If we name $CS_{a-1,a}=\int_{a-1}^{a} {m_{x} l_{a-x} dx}$, we can express the first term in the right as $CS_a - CS_{a-1,a}$, and finally divide by $CS_a$: 
$$ = - CS_a + CS_{a-1,a} (1-a+\kappa_{a-1,a})\\
\frac{\Delta CS_{a}}{CS_{a}} \approx - \left[1-\frac{CS_{a-1,a}}{CS_a} (1-(a-\kappa_{a-1,a}))\right] \Delta\delta$$

This means that an absolute change in infant hazard rate affects proportionally for all the age range extension (because of all the cohorts aged more than 1), except for the mean time exposure for those who born and survive in the last year $1-(a-k_{a-1,a})$, times $\frac{CS_{a-1,a}}{CS_a}$ the portion of cummulative alive daughters for the same period, weighting how important was this in total successful experience (in terms of alive descendants).

@@

A numerical approximation of both measures are done in [Applications](#Applications) section.

### Heterogeneity (first ideas, some guidance is needed)

@@

Isolating for an age *x* we can inspect the part related to z and k: $(l_{a-x})^z \, k\, f_{kz}$. 
Following @Gupta2018, one way to create a correlated bivariate distribution is assume for example that $Z = Y_0 + Y_1$ and $K = e^{(Y_0 + Y_2)}$, given the fact that the historical relation is not linear between both components (see figure \@ref(fig:plot_tfr_q0), a funtion *f* like $F_\beta=f(ln(q_{0,5}))$. The auxiliar variables $Y_0$, $Y_1$ and $Y_2$ are independet and Gamma distribuited. Then we can express, using Laplace transform properties:

@@

...

@@

...

@@

The approximation proposed for CS_a is reaalyprecise, especially for recent periods. For exaple, Guatemala improved the approximation with years due to rectangularization process in $l_x$ (figure \@ref(fig:plot_CS_aprox)).

@@

```{r CS_abs_app, fig.cap= "Effect of an absolute absolute change in mortality on child survival by maternal age of women in Latin America  for the 1950-2015 period. Goodness of fit of approximation compared to direct estimation"}

ch_abs <-  CS_change_app %>%
              ggplot() +
             geom_point(aes(x=change_obs_abs, y=change_app_abs, color=delta)) +
             geom_abline(slope=1, intercept = 0) +
             scale_x_continuous("Empirical change in CS") +
             scale_y_continuous("Approximated change in CS") +
             scale_color_continuous("Delta") +
             theme_classic()

ch_rel <-  CS_change_app %>%
              ggplot() +
             geom_point(aes(x=change_obs_rel, y=change_app_rel, color=delta)) +
             geom_abline(slope=1, intercept = 0) +
             scale_x_continuous("Empirical change in CS") +
             scale_y_continuous("Approximated change in CS") +
             scale_color_continuous("Delta") +
             theme_classic()

grid.arrange(ch_abs, ch_rel, ncol=2)

```

>>>>

Kin count estimation has a long and distinguished pedigree in mathematical demography, starting with the work of @lotka1931orphanhood on modelling orphanhood in theoretical populations across demographic regimes. To the best of our knowledge, @brass_derivation_1953 first proposed a formula to represent child survival over age. The notion of estimating the expected number of living daughters in a stable population was generalized by @goodman1974 for other kin relations (granddaughters, cousins, etc.). The "counting method" approach was further popularized in @KeyfitzMath. @bongaarts_projection_1987 used a similar approach to estimate descendants in his 'Family Status Model'. Recently @Caswell2019 reformulated these relationships in a matrix framework, extending the potential results to variance estimation, multistate models, and others.

Child survival - and it's counterpart, child death - gain new significance in the wake of the fertility transition.
Depressed fertility means that fewer children are tasked with providing key emotional, social, and financial transfers to aging parents for ever increasing periods of time.
With increasing periods of generational overlap, individuals find themselves 'sandwiched' between aging parents and young children requiring their simultaneous attention and care [@daatland_norwegian_2010].
In the context of global population aging, elderly parents without access to formal social security and pension systems are particularly reliant on these transfers to make ends meet [@smith-greenaway_maternal_2020].  
Losing an only child may become a more common experience in the context of global fertility decline, something particularly worrying for parents suddenly rendered childless.
Given the known psychological, health, and social consequences of child loss for bereaved parents and families [@Hendrickson2009;@lee_death_2014], it is surprising that the demographic processes that shape parental bereavement remain very poorly understood.
Recent development have underscored the need to understand how sudden changes in mortality affect the availability of kin. 
In the context of the global pandemic caused by the Covid-19 disease, elderly people depend on support provided by their relatives, especially during periods of lockdown and in the absence of governmental support mechanisms.<!-- Maybe here are repeated --> Elderly parents may be at a higher risk of losing the key support provided by their adult children given the known age-gradient in the Covid-19 case fatality rate.

How do changes in mortality affect the availability of offspring over age from the point of view of a prospective mother? How can we characterize the timing of offspring mortality as experienced by a mother from a formal demographic perspective? How this would work in an heterogeneity population?
In the remainder of the paper, we aim to formalize the relationship population-level changes in mortality rates and perceived change in the lived experience of child death from the perspective of a mother. Our approach is relevant for understanding the impact of overall and age-specific excess mortality on the resilience or otherwise of kinship networks for the vulnerable elderly population.
We consider the case when these changes are additive and mutliplicative. In this paper, we uncover formal relationships to answer these questions and provide numerical applications using data from countries in Latin America during the second half of the twentieth century. See the [Applications](#Applications) section for more details on the source, calculation procedures and main numerical results. We used period measures, in an stable context, so the objetive is give moment measures. A more sophisticated model using the appropriate age-specific cohort rates in the subscripts has been proposed by @alburez_gutierrez_womens_2019.

@@

As an initial step, we seek an intuitive understanding of the previous relation. Building on work by @KeyfitzMath for the probability of a living mother, we can show which features of the child survival function explain the expected number of children. 
For this we use an approximate of $l_x$ using Taylor´s theorem until second order around the mean age of childbearing $\kappa$:

@@

We find that, seen from the perspective of a mother, child survival mainly depends on the cummulative fertility function and the survival of daughters from birth to $a-\kappa$. Also that the approximation is affected (negativally) by how disperse is the fertility by age (variance $\sigma^2$) and the survival curvature around (with negative sign), in an age range where $l_x$ is tipycally very flat (range 20-40 years old) in transitioned populations^[This approximation is useful also for get a dimension changes in the fertility average age $\kappa$, keeping the level: taking logs in $F_a \, l_{a-\kappa}$ and deriving we get $\frac{\Delta CS_a}{CS_a} \approx - \mu_{a-\kappa} \Delta\kappa$].

@@

$$\frac{dCS^{\delta}}{d\delta} = -a\int_{\alpha}^{a}m_{x}{l_{a-x} dx} + \int_{\alpha}^{a} x m_{x} {l_{a-x} dx}\\
= -a \, CS_{a}  + \int_{\alpha}^{a} {x  \, m_{x} l_{a-x} dx}$$

@@

The expected change in descendants survival is inversely proportional to the difference between maternal age $a$ and the mean age of the mother at the birth of her surviving daughters $k_a$. The magnitude of the change depends negatively on the age distribution of the surviving offspring (the younger the fertility bigger the impact), which is intuitive considering that older descendants experiences longer periods of exposure to risk.

That an equal change in mortality could happens at all ages is unlikelihood. This measure could be much related a change in first ages (child and youth) or adolescent ones. If the change only happens in the age range $\left[0;1\right)$, infant mortality, we can inspect the effect in child survival $CS_{a}^{\delta_0}$ starting with splitting the integral between those daughters who were exposed all the range and those who do not (now generalizing for $\alpha$ positive and $a>1$ for simplicity):

@@

$$\frac{dCS^{\delta_0}}{d\delta_0} =-\int_{0}^{a-1} m_{x} l_{a-x} dx - a \int_{a-1}^{a} {m_{x} l_{a-x} dx} + \int_{a-1}^{a} {x \, m_{x} l_{a-x} dx}$$
If we name $CS_{a-1,a}=\int_{a-1}^{a} {m_{x} l_{a-x} dx}$, we can express the first term in the right as $CS_a - CS_{a-1,a}$, and recognizing that last term could be expressed as  $\kappa_{a-1,a} CS_{a-1,a}$, we get:

$$ \frac{dCS^{\delta_0}}{d\delta_0} = - CS_a + CS_{a-1,a} (1-a+\kappa_{a-1,a})$$
Where the factor $1-(a-\kappa_{a-1,a})$, which lies between 0 and 1, is equal to the average portion of time spent with living daughters durings age *a*. Assuming uniform distribution of those births, discretizing the change and dividing by $CS_a$ we get the final expression for change in infant mortality:

$$\frac{\Delta CS_{a}^{\delta_0}}{CS_{a}} \approx - \left[1-\frac{CS_{a-1,a}}{CS_a}\frac{1}{2}\right] \Delta\delta$$

This means that an absolute change in infant hazard rate affects proportionally for all the age range extension (cohorts aged more than 1), except for "half" the portion $\frac{CS_{a-1,a}}{CS_a}$ on how important was this las age in total "successful" experience (in terms of alive descendants). If *a* is near *\beta* then $\frac{\Delta CS_{a}^{\delta_0}}{CS_{a}} \approx -\Delta\delta$.

@@

Following this, in populations with high infant mortality $MAD_{a-x}<1$ for all *x*, so $MAL_a\approx\kappa_a+f_0$, being $f_0$ the average time spent for those child that dies in their first year of life. A numerical approximation of both measures are done in [Applications](#Applications) section.

### Heterogeneity (first ideas)

<!--quiero ver la correlacion entre fecundidad y mortalidad infantil. 
la proporción de hijas muertas es la proporción de hermanas muertas-->

@@

Isolating for an age *x* we can inspect the part related to z and k: $(l_{a-x})^z \, k\, f_{kz}$. 
Following @Gupta2018, one way to create a correlated bivariate distribution is assume for example that $Z = Y_0 + Y_1$ and $K = e^{(Y_0 + Y_2)}$, given the fact that the historical relation is not linear between both components (see figure \@ref(fig:plot_tfr_q0), a funtion *f* like $F_\beta=f(ln(q_{0,5}))$. The auxiliar variables $Y_0$, $Y_1$ and $Y_2$ are independet and Gamma distribuited. Then we can express, using Laplace transform properties:

@@

To continue...

<!--

@@

-->

@@

The approximation proposed for CS_a is realy precise, especially for recent periods. For example, Guatemala improved the approximation with years due to rectangularization process in $l_x$ (figure \@ref(fig:plot_CS_aprox)).

@@

```{r CS_abs_app, fig.cap= "Effect of absolute and relative change in mortality on child survival by maternal age of women in Latin America for the 1950-2015 period. Goodness of fit of approximation compared to direct estimation"}

ch_abs <-  CS_change_app %>%
              ggplot() +
             geom_point(aes(x=change_obs_abs, y=change_app_abs, color=delta)) +
             geom_abline(slope=1, intercept = 0) +
             scale_x_continuous("Empirical") +
             scale_y_continuous("Approximated") +
             scale_color_continuous("Delta") +
             ggtitle("Absolute change")+
             theme_classic()

ch_rel <-  CS_change_app %>%
              ggplot() +
             geom_point(aes(x=change_obs_rel, y=change_app_rel, color=delta)) +
             geom_abline(slope=1, intercept = 0) +
             scale_x_continuous("Empirical") +
             scale_y_continuous("Approximated") +
             scale_color_continuous("Delta") +
             ggtitle("Relative change") +
             theme_classic()

grid.arrange(ch_abs, ch_rel, ncol=2)
```
